Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-04 paper.tex
lots

The other headline feature of HoTT's generalised equality is univalence: put
simply, it allows us to treat isomorphic types as equal, giving us in the
constructive world access to a technique that is commonplace classically.
It turns out that finite types have a lot to say about isomorphisms, and we will
rely heavily on this aspect of HoTT to work with them.
\subsection{Contributions}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.

In section~\ref{strong-finiteness}, we will explore the ``strong'' notions of
finiteness (i.e. those at least as strong as Kuratowski finiteness
\cite{kuratowskiNotionEnsembleFini1920}), with a special focus on cardinal
finiteness.
We will delineate some subtly different finiteness predicates, and prove the
relationship between each.
We will prove, then, that every discrete Kuratowski finite type is cardinal
finite, and subsequently will show that discrete Kuratowski finite sets form a
topos.

In section~\ref{species-and-containers}, we will redefine our finiteness
predicates using \emph{containers}, and prove that the new definitions are
equivalent to the old.
Using this, we will prove much stronger relationships between our finiteness
predicates and relation-based definitions of finiteness.
Finally, We will show how the species-container duality falls out naturally from
these new definitions.

In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.
\section{Strong Finiteness} \label{strong-finiteness}
\begin{figure}[h]
  \centering
  \begin{tikzcd}[cramped, row sep=small]
    &
    \ar[ddddd, dash, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Ordered} &
    \ar[ddddd, dash, dashed, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Unordered}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    &
    &
    \text{Split Enumerable}
      \; \ref{split-enumerable}
      \ar[d, xshift=-1ex, bend right=30]
      &
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Discrete} &
    &
    \text{Manifest Bishop}
      \ar[u, xshift=1ex, bend right=30]
      \ar[rr, end anchor=south west, start anchor=south east, bend right=10, crossing over]
      \; \ref{manifest-bishop}
      &
    &
    \text{Cardinal}
      \; \ref{cardinal}
      \ar[ll, start anchor=north west, end anchor=north east, bend right=10, "\text{linOrd}" description]
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, dashed, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Non Discrete} &
    {} &
    \text{Manifest Enumerable}
      \; \ref{manifest-enumerability}
      \ar[from=uu, xshift=-1ex, bend right=30, crossing over]
      \ar[uu, bend right=30,  "\text{Discrete}" description]
      \ar[rr, start anchor=south east, end anchor=south west, bend right=10, crossing over]
      &
    {}
    &
    \text{Kuratowski}
      \; \ref{kuratowski}
      \ar[ll, start anchor=north west, end anchor=north east, bend right=10,  "\text{linOrd}" description]
      \ar[uu, xshift=1ex, bend right=30,  "\text{Discrete}" description]
      \ar[from=uu, xshift=-1ex, bend right=30, crossing over]
  \end{tikzcd}
  \caption{Classification of the strong definitions of finiteness, according to
    whether they are discrete (imply decidable equality) and whether they induce
    a linear order.}
  \label{finite-classification}
\end{figure}

In this section, we will explore five predicates for finiteness, classified in
figure~\ref{finite-classification}.
Of these, manifest enumerability is new, to our knowledge.

Figure~\ref{finite-classification} organises the predicates according to
their ``strength''; i.e. how much information they provide about a conforming
type.
For instance, a proof that some type \(A\) is manifestly Bishop finite (the
strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
us that \(A\) is discrete (has decidable equality), and gives us a linear order
on the type.
A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
features: indeed, we will see examples of Kuratowski finite types which are not
even sets, never mind discrete ones.

In the following sections, we will go through each of the predicates, proving
how to weaken each (i.e. we will provide a proof that every cardinally finite
type is Kuratowski finite), and how to strengthen them, given the required
property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos. \todo[noline]{reference for this}
More specifically, we show that cardinal finite sets form a topos, that
decidable Kuratowski finite sets are equivalent in strength to cardinal finite
sets, and carry the proof over.
This result relies on proofs on each of the other finiteness predicates.
\subsection{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \sum_{(\mathit{xs} : \textbf{List} A)} , \prod_{(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{definition}

@@

\begin{definition}[List Membership] \label{list-membership}
  \begin{equation}
    x \in \mathit{xs} = \sum_{(i : \mathbf{Fin}(\text{length}(\mathit{xs})))} , \mathit{xs} \ind i \equiv x
  \end{equation}
  Here, \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\),
  and \(\ind\) is an indexing function.
  In other words, to say that some item \(x\) is in some list \(\mathit{xs}\),
  we say that there is an index into \(\mathit{xs}\) which selects an element of
  \(\mathit{xs}\) which is equivalent to \(x\).
\end{definition}

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsubsection{Decidable Equality}
We will now prove that split enumerability implies decidable equality.
We first need to prove a lemma regarding surjections to split enumerable types.
\begin{lemma} \label{split-enum-surj}
  Any split enumerable type has a split surjection from a finite prefix of the
  natural numbers.
\end{lemma}
\begin{proof} 
  Let \(E\) be a proof of type \(\mathcal{E}!(A)\) for some type \(A\), and
  \(n\) be the length of the support list of \(E\).
  Then let \(f : \textbf{Fin}(n) \rightarrow A\) be defined as the indexing
  function into the support list of \(E\).

  We then prove this function to be a split surjection by providing a fiber over
  all of the elements of its domain: this fiber is the second component of \(E\).

  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{to-surj}
\end{proof}

Because of careful arrangement of previous definitions, this proof is only three
lines of Agda.
It is not difficult to derive decidable equality from this fact.
\begin{lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{lemma}
\begin{proof}
  We use a corollary, proven in the appendix, that if there is a
  split-surjection from \(A\) to \(B\), and \(A\) is discrete, then \(B\) is
  also discrete.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{discrete}
\end{proof}

This also tells us that any split enumerable type must be a set.
\begin{lemma}
  Any split enumerable type is a set.
\end{lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}

Though it's not strictly needed for the rest of the construction, it is
instructive to also prove inverse regarding surjections; i.e. that we can
derive \(\mathcal{E}!\) from a surjection from another finite type.
\begin{lemma} \label{split-enum-from-surj}
  Any type with a split surjection from another split enumerable type is itself
  split enumerable.
\end{lemma}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\), and let \(f\)
  be a surjection from \(A\) to \(B\).
  The support list for the proof of split enumerability over \(B\) can be
  obtained by mapping the surjective function over the support list of \(E\).

  To obtain the cover proof, let \(f^{-1}\) be the inverse function of \(f\)
  (which we can obtain because it is a split-surjection).
  Using the cover proof of \(E\), we can show that, for all \(y : B\), \(f^{-1}(y)\)
  is in the support list for \(E\).
  Using the following lemma (proven in the appendix):
  \begin{equation}
    x \in xs \rightarrow f (x) \in \text{map} \; f \; \mathit{xs}
  \end{equation}
  We obtain a proof that:
  \begin{equation}
    \forall (y : B). f (f^{-1}(y)) \in \text{map} \; f \; \mathit{xs}
  \end{equation}
  And, since \(f (f^{-1}(y)) \equiv y\) (by the definition of a surjection), we
  have proven our goal.

  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{from-surj}
\end{proof}
\subsubsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{lemma}
  \(\bot\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{lemma}
  \(\top\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{theorem}
  Split-enumerability is closed under \(\sum\).
\end{theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

  The cover proof follows in a simple, if tedious, way.
\end{proof}
\begin{lemma}
  Split-enumerability is closed under Cartesian product (non-dependent product).
\end{lemma} 
\begin{proof}
  Since non-dependent products are simply a special case of the \(\sum\) type,
  we can reuse the proof of closure over \(\sum\) here.
\end{proof}
\begin{lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum).
\end{lemma}
\begin{proof}
  Again, this can be derived from closure of the \(\sum\) type.
  Disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
The final case can be derived from the previous closures; we give it here anyway
for clarity.
\begin{lemma} \label{fin-split-enum}
  Any finite prefix of the natural numbers (what we have been calling
  \(\textbf{Fin}\)) is split enumerable.
\end{lemma}
\begin{proof}
  For the support list, we will make use of the function tabulate, which returns
  a list consisting of the application of \(f\) to every \(\textbf{Fin}(n)\),
  given \(n\).
  \begin{equation}
    \text{tabulate}(n , f) = \left[ f(0) , f(1) \dots f(n - 1) \right]
  \end{equation}
  In Agda:
  \ExecuteMetaData[agda/Data/List.tex]{tabulate}
  Tabulation with the identity function gives us our cover list.

  The cover proof is obviously true given the definition of tabulate above; in
  Agda it requires a little more work to show.
  \ExecuteMetaData[agda/Data/List/Membership.tex]{fin-in-tabulate}
\end{proof}
\subsection{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
It is defined in terms of our previous list membership term
(definition~\ref{list-membership}).
\begin{definition}[Unique List Membership]
  We say an item \(x\) is ``uniquely in'' some list \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsubsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{lemma} \label{bishop-equiv}
  Any manifestly Bishop finite type has an equivalence with a finite prefix of
  the natural numbers.
\end{lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}. \todo[noline]{Need this in appendix also}
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.

  First, we define the map as the indexing function into the support list.
  Proving this map is a contraction is precisely the cover proof.

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{equiv}
\end{proof}

As we can see, our list-based definition of manifest Bishop finiteness is really
an equivalence with \(\mathbf{Fin}\) in disguise.
This will be the key component in establishing closure under exponentials.

We would like to draw attention to the ``power-to-weight ratio'' of this proof.
We have mentioned how important the equivalence will be for our later proofs: in
general, the ability to transport via univalence opens up many opportunities

In Cubical Type Theory, equivalences are much more useful than their counterpart
(isomorphisms) in Martin-Löf  Type Theory: arbitrary proofs and functions can be
\emph{transported} along equivalences, giving a constructive justification for
an intuitive and useful technique which is commonplace classically.
The proof itself, however, is three lines long: thanks to careful definitions,
the fact that manifest Bishop finiteness is the ``same'' as equivalences to
\(\mathbf{Fin}\) becomes obviously true.
Later we will see, by defining lists as containers, a proof that manifest Bishop
finiteness is equivalent (in the formal sense) to equivalences with
\(\mathbf{Fin}\). \todo[noline]{This needs to be added to the containers
  section.}
\todo[noline]{More on this? }
\subsubsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{lemma}
  Any manifest Bishop finite set is split enumerable.
\end{lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-to-split-enum}
\end{proof}
\begin{theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

@@

  Full details of the formal proof are provided in the appendix.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsubsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).

>>>>

The other headline feature of HoTT's generalised equality is univalence: put
simply, it allows us to treat isomorphic types as equal, giving us in the
constructive world access to a technique that is commonplace classically.
It turns out that finite types have a lot to say about isomorphisms, and we will
rely heavily on this aspect of HoTT to work with them.
\subsection{Contributions}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.
\subsubsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

\input{figures/finite-classification}
Figure~\ref{finite-classification} organises the predicates according to
their ``strength''; i.e. how much information they provide about a conforming
type.
For instance, a proof that some type \(A\) is manifestly Bishop finite (the
strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
us that \(A\) is discrete (has decidable equality), and gives us a linear order
on the type.
A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
features: indeed, we will see examples of Kuratowski finite types which are not
even sets, never mind discrete ones.

We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.

We will---through the use of containers
\cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
these predicates have with the usual function relations i.e. we will show that a
proof of manifest enumerability is precisely equivalent to a surjection from a
finite prefix of the natural numbers.

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos. \todo[noline]{reference for this}
More specifically, we show that cardinal finite sets form a topos, that
decidable Kuratowski finite sets are equivalent in strength to cardinal finite
sets, and carry the proof over.
This result relies on proofs on each of the other finiteness predicates.
\subsubsection{Species}
\todo[inline]{Redo this next paragraph}
In section~\ref{species-and-containers}, we will redefine our finiteness
predicates using \emph{containers}, and prove that the new definitions are
equivalent to the old.
Using this, we will prove much stronger relationships between our finiteness
predicates and relation-based definitions of finiteness.
Finally, We will show how the species-container duality falls out naturally from
these new definitions.

\subsubsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

\subsubsection{Practical Uses of Finiteness}
Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{definition}

@@

\begin{definition}[List Membership] \label{list-membership}
  \begin{equation}
    x \in \mathit{xs} = \sum_{(i : \mathbf{Fin}(\text{length}(\mathit{xs})))} , \mathit{xs} \ind i \equiv x
  \end{equation}
  Here, \(\ind\) is a safe indexing function.
  In other words, to say that some item \(x\) is in some list \(\mathit{xs}\),
  we say that there is an index into \(\mathit{xs}\) which selects an element of
  \(\mathit{xs}\) which is equivalent to \(x\).
\end{definition}
\begin{definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= 1 + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{definition}

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsection{Proving Equivalence with Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
Here, we will formalise this equivalence, by first proving an equivalence with a
container-based definition of split enumerability.

Containers \cite{abbottContainersConstructingStrictly2005} give us a general way
to construct functors from generic building-blocks.
\begin{definition}[Container]
  A container is a dependent pair of shapes, and positions indexed by those
  shapes.
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
\end{definition}

For instance, the shape of lists is given by \(\mathbb{N}\), and the positions are then
\(\mathbf{Fin}\)s indexed by that \(\mathbb{N}\).
If we swap out wherever we used ``list'' in the definition of split
enumerability with this container-based list, we get an equivalent finiteness
predicate.
\begin{lemma} \label{container-split-enum-equiv}
  The container-based definition of split enumerability is equivalent to the
  inductive type version.
  \begin{equation}
    \sum_{(\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket A)} , \prod_{(x : A)} , x \in xs \simeq
    \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
  Membership for container-defined lists is a fiber.
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

This version of split enumerability is equivalent to a split surjection from a
finite prefix of the natural numbers. 
\begin{lemma} \label{container-split-enum-surj}
  Container-defined split enumerability is equivalent to a split surjection from
  some \(\mathbf{Fin}\).
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \mathbf{Fin}(n) \twoheadrightarrow ! A
  \end{equation}
\end{lemma}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
This lets us prove our overall goal for this section.
\begin{theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
\end{theorem}
\begin{proof}
  This follows from lemmas \ref{container-split-enum-equiv} and
  \ref{container-split-enum-surj}.
\end{proof}
\subsection{Decidable Equality}
\begin{lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{lemma}
\begin{proof}
  We use a corollary, proven in the appendix, that if there is a
  split-surjection from \(A\) to \(B\), and \(A\) is discrete, then \(B\) is
  also discrete.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{discrete}
  \todo[inline]{Rework this proof to rely on the split surjection equivalence
    from before}
\end{proof}

This also tells us that any split enumerable type must be a set.
\begin{lemma}
  Any split enumerable type is a set.
\end{lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{lemma}
  \(\bot\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{lemma}
  \(\top\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{theorem}
  Split-enumerability is closed under \(\sum\).
\end{theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

  The cover proof follows in a simple, if tedious, way.
\end{proof}
\begin{lemma}
  Split-enumerability is closed under Cartesian product (non-dependent product).
\end{lemma} 
\begin{proof}
  Since non-dependent products are simply a special case of the \(\sum\) type,
  we can reuse the proof of closure over \(\sum\) here.
\end{proof}
\begin{lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum).
\end{lemma}
\begin{proof}
  Again, this can be derived from closure of the \(\sum\) type.
  Disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
It is defined in terms of our previous list membership term
(definition~\ref{list-membership}).
\begin{definition}[Unique List Membership]
  We say an item \(x\) is ``uniquely in'' some list \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  \todo[inline]{Finish prose for this proof}
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{lemma}
  Any manifest Bishop finite set is split enumerable.
\end{lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-to-split-enum}
\end{proof}
\begin{theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

@@

  Full details of the formal proof are provided in the appendix.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).

  Formally, given a type \(A\) and a type family \(U\) on \(A\), when \(A\) is
  manifestly Bishop finite:
  \begin{equation}
    \mathcal{B}\left( A \right)
  \end{equation}
  And \(U\) is manifestly Bishop finite over all points of \(A\):
  \begin{equation}
    \prod_{(x : A)} , \mathcal{B}\left( U(x) \right)
  \end{equation}
  Then we have the following:
  \begin{equation}
    \mathcal{B}\left(\prod_{(x : A)} , U(x)\right)
  \end{equation}
\end{theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.
